A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics

### 2019-03-14 pkg/colorspace/paper/Toolbox/article.Rmd
COW feedback integrated
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@486 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

  General strategies for three types of palettes are provided:
  (1)\ Qualitative for coding categorical information, i.e., where no particular
  ordering of categories is available. (2)\ Sequential for coding ordered/numeric information, i.e.,
  going from high to low (or vice versa). (3)\ Diverging for coding ordered/numeric
  information around a central neutral value, i.e., where colors diverge from
  neutral to two extremes.

  To aid selection and application of these palettes the package provides scales
  for use with \pkg{ggplot2}, \pkg{shiny} (and \pkg{tcltk}) apps for interactive
  exploration, visualizations of palette properties,
  accompanying manipulation utilities (like desaturation and lighten/darken), and
  emulation of color vision deficiencies.
keywords:
  # at least one keyword, comma-separated and in sentence case
  formatted: [color, palette, HCL, RGB, hue, color vision deficiency, "\\proglang{R}"]
  plain:     [color, palette, HCL, RGB, hue, color vision deficiency, R]
preamble: >
  \usepackage{thumbpdf,lmodern}
  \usepackage{booktabs,amsmath,longtable,supertabular,framed}
  \shortcites{color:colorspace,color:ggplot2pkg}
  \newcommand{\tex}[1]{#1}
output: rticles::jss_article
bibliography: ../../vignettes/color.bib
---

@@

To address these problems, many improved color palettes with better perceptual
properties have been suggested in the literature over the last one or two decades
[@color:Harrower+Brewer:2003; @color:Zeileis+Hornik+Murrell:2009; @color:Smith+VanDerWalt:2015; @color:CARTO; @color:Crameri:2018].
Many systems for systems for statistical and scientific computing provide
infrastructure for such color palettes, e.g., for \proglang{R} [@color:R] the
list of useful packages encompasses:
\pkg{RColorBrewer} [@color:RColorBrewer],
\pkg{viridis} [@color:viridis],
\pkg{rcartocolor} [@color:rcartocolor],
\pkg{wesanderson} [@color:wesanderson],
or \pkg{scico} [@color:scico] among many others.
Furthermore, packages like \pkg{pals} [@color:pals] and \pkg{paletteer} [@color:paletteer]
collect many of the proposed palettes in combination with a unified interface.
Most of these palettes, however, are pre-existing palettes, stored as a
limited set of colors and interpolated as necessary. And even if specific
algorithms have been used in the initial construction of the palettes, these
are often not reflected in the software implementations.

```{r hcl-properties, echo = FALSE, fig.width = 4, fig.height = 2.2, out.width = "0.5\\textwidth", fig.cap = "Axes of the HCL color space. Top: Hue $H$ changes from 0 (red) via 60 (yellow), etc.\\ to 300 (purple) with fixed $C = 60$ and $L = 65$. Center: Chroma $C$ changes from 0 (gray) to 100 (colorful) with fixed $H = 0$ (red) and $L = 65$. Bottom: Luminance $L$ changes from 90 (light) to 25 (dark) with fixed $H = 260$ (blue) and $C = 25$ (low, close to gray)."}
swatchplot(
  "Hue\n(Type of color)"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
  "Chroma\n(Colorfulness)"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
  "Luminance\n(Brightness)" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
  nrow = 3, line = 5.2, off = 0
)
```

The \pkg{colorspace} package [@color:colorspace] adopts a somewhat different
approach, utiliting the perceptually-based HCL (Hue--Chroma--Luminance) color space
[@color:Wiki+HCL]. The three axes of the HCL space match those of the human visual
system very well: Hue (= type of color, dominant wavelength),
chroma (= colorfulness), luminance (= brightness), see Figure\ \ref{fig:hcl-properties}.
Thus, based on this color model the \pkg{colorspace} package can derive general and adaptable
strategies for color palettes; manipulate individual colors and color palettes;
and assess and visualize the properties of color palettes (beyond simple color swatches).
Specifically, \pkg{colorspace} provides three types of palettes based on the HCL model:

@@

A broad collection of prespecified palettes are shipped in the package, existing
palettes can be easily tweaked, and new or adapted palettes registered. The prespecified
palettes include suitable HCL color choices that closely approximate of most palettes
from packages \pkg{RColorBrewer}, \pkg{rcartocolor}, and \pkg{viridis} by using only a
small set of hue, chroma, and luminance parameters.

To aid choice and application of these palettes there are
(a) scales for use with \pkg{ggplot2} [@color:ggplot2],
(b) \pkg{shiny} [@color:shiny] [and \pkg{tcltk}, @color:R] apps for interactive exploration,
(c) visualizations of palette properties, and
(d) accompanying manipulation utilities (like desaturation, lighten/darken, and emulation of color vision deficiencies).

The remainder of the manuscript is organized as follows:
Section\ \ref{sec:tour} gives a first overview of the package's "look & feel" and the general workflow.
Section\ \ref{sec:hcl_palettes} introduces the extensible collection of HCL-based palettes along with their construction details.
Section\ \ref{sec:palette_visualization} presents the toolbox for palette visualization and assessment.
Section\ \ref{sec:color_vision_deficiency} discusses the implemented techniques for color vision deficiency emulation
that help assess the suitability of colors for colorblind viewers.
Section\ \ref{sec:hclwizard} briefly highlights the interactive color apps from the package.
Finally, some further color manipulation utilities are highlighted in Section\ \ref{sec:manipulation_utilities}
before Section\ \ref{sec:summary} concludes the manuscript.

@@

```{r hcl-palettes, fig.height = 6.75, fig.width = 12, fig.cap = "Overview of available predefined palettes in \\pkg{colorspace}."}
library("colorspace")
hcl_palettes(plot = TRUE)
```

A suitable vector of colors can be easily computed by specifying the desired number
of colors and palette name (see Figure\ \ref{fig:hcl-palettes} for possible palette names), e.g.,

@@

The functions `sequential_hcl()`, and `diverging_hcl()` work analogously. Additionally,
a palette's hue/chroma/luminance parameters can be modified, thus allowing to easily customize
each palette. Moreover, the `choose_palette()`/`hclwizard()` app provides convenient user
interfaces to perform palette customization interactively. Finally, even more flexible diverging
HCL palettes are provided by `divergingx_hcl()`.

@@

As another example for a sequential palette a spine plot is created (see the right
panel of Figure\ \ref{fig:eustockmarkets-titanic}),
displaying the proportion of Titanic passengers that survived per class.
The `"Purples 3"` palette is used which is quite similar to the \pkg{ColorBrewer.org}
[@color:Harrower+Brewer:2003] palette `"Purples"`. Here, only two colors are
employed, yielding a dark purple and light gray.

@@

To plug the HCL color palettes into \pkg{ggplot2} graphics [@color:ggplot2; @color:ggplot2pkg]
suitable discrete and/or continuous \pkg{gglot2} color scales are provided.
The scales are called via the scheme

@@

To illustrate their usage two simple examples are shown using the qualitative `"Dark 3"`
and sequential `"Purples 3"` palettes that were also employed above
(see Figure\ \ref{fig:iris-diamonds}). For the first example, semi-transparent
shaded densities of the sepal length from the `iris` data are shown, grouped by species
(see the left panel of Figure\ \ref{fig:iris-diamonds}).

@@

In Figure\ \ref{fig:allplots-sequential}, a heatmap (based on the well-known Maunga Whau volcano data) is used as a typical
application for a sequential palette. The elevation of the volcano is brought out clearly,
focusing with the dark colors on the higher elevations.
The other two displays show that hue is constant in the palette while luminance and chroma vary.
Luminance increases monotonically from dark to light (as required for a proper sequential palette).
Chroma is triangular-shaped which allows to better distinguish the middle colors in the palette
when compared to a monotonic chroma trajectory.

@@

The color spaces, implemented in \pkg{colorspace}, along with corresponding \proglang{S}4 classes
and eponymous class constructors are:

* `RGB()` for the classic Red-Green-Blue color model that mixes three primary colors
  with different intensities to obtain a spectrum of colors. The advantage of this
  color model is (or was) that it corresponded to how computer and TV screens generated
  colors, hence it was widely adopted and still is the basis for color specifications
  in many systems. For example, the hex color codes are employed in HTML but also in \proglang{R}.
  However, the RGB model also has some important drawbacks: It does not take into
  account the output device properties, it is not perceptually uniform (a unit step
  within RGB does not produce a constant perceptual change in color), and it is
  unintuitive for humans to specify colors (say brown or pink) in this space.
  See @color:Wiki+RGB for more details.
* `sRGB()` addresses the issue of device dependency by adopting a so-called gamma
  correction. Therefore, the gamma-corrected standard RGB (sRGB), as opposed to the
  linearized RGB above, is a good model for specifying colors in software and for
  hardware. But it is still unintuitive for humans to work directly with this.
  Therefore, sRGB is a good place to end up in a color space manipulation but
  it is not a good place to start. See @color:Wiki+sRGB for more details.
* `HSV()` is a simple transformation of the (s)RGB space that tries to capture the
  perceptual axes: _hue_ (dominant wavelength, the type of color), _saturation_
  (colorfulness), and _value_ (brightness, i.e., light vs.\ dark). Unfortunately,
  the three axes in the HSV model are confounded so that, e.g., brightness changes
  dramaticaly with hue.
  See @color:Wiki+HSV for more details.
* `HLS()` (Hue-Lightness-Saturation) is another transformation of (s)RGB that tries
  to capture the perceptual axes. It does a somewhat better job but the dimensions
  are still strongly confounded.
  See @color:Wiki+HSV for more details.
* `XYZ()` was established by the CIE (Commission Internationale de l'Eclairage)
  based on psychophysical experiments with human subjects. It provides a unique triplet of
  XYZ values, coding the standard observer's perception of the color. It is
  device-independent but it is not perceptually uniform and the XYZ coordinates
  have no intuitive meaning.
  See @color:Wiki+CIEXYZ for more details.
* `LUV()` and `LAB()` were therefore proposed by the CIE as perceptually uniform
  color spaces where the former is typically preferred for emissive technologies
  (such as screens and monitors) whereas the latter is usually preferred when
  working with dyes and pigments. However, the three axes of these two spaces
  still do not correspond to human perceptual axes.
  See @color:Wiki+CIELUV; @color:Wiki+CIELAB for more details.
* `polarLUV()` and `polarLAB()` therefore take polar coordinates in the UV plane
  and AB plane, respectively. Specifically, the polar coordinates of the LUV
  model are known as the HCL (Hue-Chroma-Luminance) model. These capture the
  human perceptual axes very well without confounding effects as in the HSV or
  HLS approaches. (More details follow below.)

All \proglang{S}4 classes for color spaces inherit from a virtual class `color` which are
internally always represented by matrices with three columns (corresponding to
the different three dimensions).

Note that since the inception of the color space conversion tools within \pkg{colorspace}'s
\proglang{C} code by @color:Ihaka:2003 other \proglang{R} tools for this purpose became available.
Notably, base \proglang{R} meanwhile provides `grDevices::convertColor()`
[computed in high-level \proglang{R}, @color:R] and
`farver::convert_colour()` [@color:farver] is based on a \proglang{C++} library. For many basic color
conversion purposes the \pkg{colorspace} package and these alternatives are essentially equally suitable,
though [see the discussion in @color:convertColor]. For more complex conversions, including
different chromatic adaptation algorithms, a more comprehensive color science approach
is implemented in the \proglang{R} package \pkg{colorscience} [@color:colorscience].
Finally, base \proglang{R} also provides
`grDevices::hcl()` for mapping HCL representations to hex codes.

@@

As motivated in the previous section,
the HCL space is particularly useful for specifying individual colors and color palettes as its three axes match those
of the human visual system very well. Therefore, the \pkg{colorspace} package provides
three palette functions based on the HCL model: `qualitative_hcl()`, `sequential_hcl()`,
and `diverging_hcl()`. Their construction principles are exemplified in
Figure\ \ref{fig:hcl-palettes-principles} and explained in more detail below. The desaturated
palettes in the second row of Figure\ \ref{fig:hcl-palettes-principles}
bring out clearly that luminance differences (= light-dark contrasts)
are crucial for sequential and diverging palettes while qualitative palettes
are balanced at the same luminance.

@@

As suggested by @color:Ihaka:2003 `qualitative_hcl()` distinguishes the underlying categories by a sequence of
hues while keeping both chroma and luminance constant to give each color in the
resulting palette the same perceptual weight. Thus, `h` should be a pair of
hues (or equivalently `h1` and `h2` can be used) with the starting and
ending hue of the palette. Then, an equidistant sequence between these hues is
employed, by default spanning the full color wheel (i.e., the full 360 degrees).
Chroma `c` (or equivalently `c1`) and luminance `l` (or equivalently
`l1`) are constants.

@@

When qualtitative palettes are employed for shading areas in statistical displays
(e.g., in bar plots, pie charts, or regions in maps), then it is typically
less distracting to use lighter colors (with moderate chroma and high luminance)
such as `"Pastel 1"` or `"Set 3"`. When coloring points or lines then more flashy
colors (with high chroma) are often required: On a white background a moderate
luminance as in `"Dark 2"` or `"Dark 3"` usually works better while on a black/dark
background the luminance should be higher as in `"Set 2"`.
Some examples with demo graphics are provided in Section\ \ref{sec:palette_visualization}.

@@

As suggested by @color:Zeileis+Hornik+Murrell:2009 `sequential_hcl()` codes the underlying numeric values by a monotonic sequence
of increasing (or decreasing) luminance. Thus, the function's `l` argument should provide
a vector of length 2 with starting and ending luminance (equivalently, `l1` and
`l2` can be used). Without chroma (i.e., `c = 0`), this simply corresponds
to a grayscale palette like `gray.colors()`, see `"Grays"` and `"Light Grays"`
in Figure\ \ref{fig:hcl-palettes-sequentials}.

@@

For distinguishing colors in the center of the palette, two strategies can be employed:
(a) Hue can be varied as well by specifying an interval of hues in `h` (or beginning
hue `h1` and ending hue `h2`). More details are provided in the next section.
(b) Instead of a decreasing chroma a triangular chroma trajectory can be employed
from `c1` over `cmax` to `c2` (or equivalently a vector `c` of
length 3). This yields high-chroma colors in the middle of the palette that are
more easily distinguished from the dark and light extremes. See `"Blues 3"`, `"Purples 3"`,
`"Reds 3"`, and `"Greens 3"` in Figure\ \ref{fig:hcl-palettes-sequentials}.

Instead of employing linear trajectories in the chroma or luminance coordinates,
some palettes employ a power transformation of the chroma and/or luminance
trajectory. Either a vector `power` of length 2 or separate `p1` (for chroma)
and `p2` (for luminance) can be specified. If missing, the latter defaults to the
former.

@@

All except the last palette are inspired by the \pkg{ColorBrewer.org} palettes with
the same base name [@color:Harrower+Brewer:2003] but restricted to a single hue only.
They are intended for a white/light background. The last palette (`"Oslo"`) is taken from
the scientific color maps of @color:Crameri:2018 and is intended for a black/dark
background and hence the order is reversed starting from a light blue (not a light gray).

@@

This uses a slight power transformation with `p1 = 1` in the blue arm of the palette but otherwise essentially
corresponds to what `cividis_hcl()` does. For convenience `divergingx_hcl(n, palette = "Cividis")` is preregistered
using the above parameters.

@@

As a somewhat more complex example, a `swatchplot()` is created with three matrices
of sequential color palettes of blues, purples, reds, and greens. For all palettes,
luminance increases monotonically to yield a proper sequential palette. However,
the hue and chroma handling is somewhat different to emphasize different parts of the
palette.

@@

While the `specplot()` function above works well for bringing out the HCL coordinates
associated with a given palette, it does not bring out how the palette fits into
the HCL space. For example, it is not so clear whether high chroma values are
close to the maximum possible for a given hue. Thus, it cannot be judged so easily
how the parameters of the hue, chroma, and luminance trajectories can be
modified to obtain another palette.

@@

```{r hcl-qualitative, fig.width = 4, fig.height = 4, dev = "png", out.width = "0.5\\textwidth", fig.cap = "Hue-chroma plane with luminance fixed at $L = 70$ along with a qualitative palette with varying hue $H$ and chroma fixed at $C = 50$."}
hclplot(qualitative_hcl(9, c = 50, l = 70))
```

@@

```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, dev = "png", fig.cap = "Luminance-chroma plane with variations of blue sequential single-hue palettes. Left: Linear chroma for $H = 260$. Center: Triangular chroma for $H = 245$. Right: Power-transformed triangular chroma for $H = 245$."}
par(mfrow = c(1, 3))
hclplot(sequential_hcl(7, h = 260, c = 80, l = c(35, 95), power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95),
  power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95),
  power = c(0.8, 1.4)))
```

@@

The workhorse function to do so is `simulate_cvd()` which can take any vector
of valid \proglang{R} colors and transform them according to a certain CVD transformation
matrix and transformation equation. The transformation matrices have been established
by @color:Machado+Oliveira+Fernandes:2009 and are provided in objects `protanomaly_cvd`,
`deutanomaly_cvd`, and `tritanomaly_cvd`.
The convenience interfaces `deutan()`, `protan()`, and `tritan()` are the high-level functions for
simulating the corresponding kind of color blindness with a given severity (calling
`simulate_cvd()` internally).

>>>>

  Using the HCL color model general strategies for three types of palettes are implemented:
  (1)\ Qualitative for coding categorical information, i.e., where no particular
  ordering of categories is available. (2)\ Sequential for coding ordered/numeric information, i.e.,
  going from high to low (or vice versa). (3)\ Diverging for coding ordered/numeric
  information around a central neutral value, i.e., where colors diverge from
  neutral to two extremes.

  To aid selection and application of these palettes the package also contains scales
  for use with \pkg{ggplot2}, \pkg{shiny} (and \pkg{tcltk}) apps for interactive
  exploration, visualizations of palette properties,
  accompanying manipulation utilities (like desaturation and lighten/darken), and
  emulation of color vision deficiencies.
keywords:
  # at least one keyword, comma-separated and in sentence case
  formatted: [color, palette, HCL, RGB, hue, color vision deficiency, "\\proglang{R}"]
  plain:     [color, palette, HCL, RGB, hue, color vision deficiency, R]
preamble: >
  \usepackage{thumbpdf,lmodern}
  \usepackage{booktabs,amsmath,longtable,supertabular,framed}
  \shortcites{color:colorspace,color:ggplot2pkg}
  \newcommand{\tex}[1]{#1}
output: rticles::jss_article
bibliography: ../../vignettes/color.bib
---

@@

To address these problems, many improved color palettes with better perceptual
properties have been receiving increasing attention in the literature
[@color:Harrower+Brewer:2003; @color:Zeileis+Hornik+Murrell:2009; @color:Smith+VanDerWalt:2015; @color:CARTO; @color:Crameri:2018].
Many systems for systems for statistical and scientific computing provide
infrastructure for such color palettes, e.g., for \proglang{R} [@color:R] the
list of useful packages encompasses:
\pkg{RColorBrewer} [@color:RColorBrewer],
\pkg{viridis} [@color:viridis],
\pkg{rcartocolor} [@color:rcartocolor],
\pkg{wesanderson} [@color:wesanderson],
or \pkg{scico} [@color:scico] among many others.
Furthermore, packages like \pkg{pals} [@color:pals] and \pkg{paletteer} [@color:paletteer]
collect many of the proposed palettes in combination with a unified interface.
Most of these palettes, however, are pre-existing palettes, stored as a
limited set of colors and interpolated as necessary. And even if specific
algorithms have been used in the initial construction of the palettes, these
are often not reflected in the software implementations.

```{r hcl-properties, echo = FALSE, fig.width = 4, fig.height = 2.4, out.width = "0.5\\textwidth", fig.cap = "Axes of the HCL color space. Top: Hue $H$ changes from 0 (red) via 75 (yellow), etc.\\ to 300 (purple) with fixed $C = 60$ and $L = 65$. Center: Chroma $C$ changes from 0 (gray) to 100 (colorful) with fixed $H = 0$ (red) and $L = 65$. Bottom: Luminance $L$ changes from 95 (light) to 15 (dark) with fixed $H = 260$ (blue) and $C = 25$ (low, close to gray)."}
par(mar = c(0.1, 6.1, 0, 0))
plot(0, 0, type = "n", xlim = c(0, 5), ylim = c(0, 3), axes = FALSE, xlab = "", ylab = "")
hcl_prop <- rbind(
  "Hue\n(Type of color)"    = hcl(0:4 * 75, 60, 65),
  "Chroma\n(Colorfulness)"  = hcl(0, 0:4 * 25, 65),
  "Luminance\n(Brightness)" = hcl(260, 25, 95 - 0:4 * 20))
mtext(rownames(hcl_prop), side = 2, at = 2:0 + 0.65, las = 1, line = 5.2, adj = 0, cex = 1)
rect(rep(0:4, each = 3), rep(2:0, 5) + 0.3, rep(1:5, each = 3), rep(3:1, 5), col = hcl_prop, border = "lightgray")
text(rep(0:4, each = 3) + 0.5, rep(2:0, 5) + 0.2, rbind(0:4 * 75, 0:4 * 25, 95 - 0:4 * 20), cex = 0.7)
```

The \pkg{colorspace} package [@color:colorspace] adopts a somewhat different
approach that gives the user direct access to the construction principles underlying
its palettes. These are based on simple trajectories in the perceptually-based
HCL (Hue--Chroma--Luminance) color space [@color:Wiki+HCL] whose axes match those of the human visual
system very well: Hue (= type of color, dominant wavelength),
chroma (= colorfulness), luminance (= brightness), see Figure\ \ref{fig:hcl-properties}.
Thus, utilizing this color model the \pkg{colorspace} package can derive general and adaptable
strategies for color palettes; manipulate individual colors and color palettes;
and assess and visualize the properties of color palettes (beyond simple color swatches).
Specifically, \pkg{colorspace} provides three types of palettes based on the HCL model:

@@

A broad collection of prespecified palettes are shipped in the package. In addition, existing
palettes can be easily tweaked and new or adapted palettes registered. The prespecified
palettes include suitable HCL color choices that closely approximate of most palettes
from packages \pkg{RColorBrewer}, \pkg{rcartocolor}, and \pkg{viridis} by using only a
small set of hue, chroma, and luminance parameters.

To aid choice and application of these palettes the package provides
(a) scales for use with \pkg{ggplot2} [@color:ggplot2],
(b) \pkg{shiny} [@color:shiny] and \pkg{tcltk} [@color:R] apps for interactive exploration,
(c) visualizations of palette properties, and
(d) accompanying manipulation utilities (like desaturation, lighten/darken, and emulation of color vision deficiencies).

The remainder of the manuscript is organized as follows:
Section\ \ref{sec:tour} gives a first overview of the package's "look & feel" and the general workflow.
Section\ \ref{sec:color_spaces} summarizes the \proglang{S}4 color space classes and methods in the package.
Section\ \ref{sec:hcl_palettes} introduces the extensible collection of HCL-based palettes along with their construction details.
Section\ \ref{sec:palette_visualization} presents the toolbox for palette visualization and assessment.
Section\ \ref{sec:color_vision_deficiency} discusses the implemented techniques for color vision deficiency emulation
that help assess the suitability of colors for colorblind viewers.
Section\ \ref{sec:hclwizard} briefly highlights the interactive color apps from the package.
Finally, some further color manipulation utilities are highlighted in Section\ \ref{sec:manipulation_utilities}
before Section\ \ref{sec:summary} concludes the manuscript.

@@

```{r hcl-palettes, fig.height = 6.75, fig.width = 12, fig.cap = "Brief overview of available predefined palettes in \\pkg{colorspace}."}
library("colorspace")
hcl_palettes(plot = TRUE)
```

A suitable vector of colors can be easily computed by specifying the desired number
of colors and the palette name (see Figure\ \ref{fig:hcl-palettes} for possible palette names), e.g.,

@@

The functions `sequential_hcl()`, and `diverging_hcl()` work analogously. Additionally,
a palette's hue/chroma/luminance parameters can be modified, thus allowing for easy customization of
each palette. Moreover, the `choose_palette()`/`hclwizard()` app provides convenient user
interfaces to perform palette customization interactively. Finally, even more flexible diverging
HCL palettes are provided by `divergingx_hcl()`.

@@

As another example for a sequential palette, we demonstrate how to create a spine plot (see the right
panel of Figure\ \ref{fig:eustockmarkets-titanic})
displaying the proportion of Titanic passengers that survived per class.
The `"Purples 3"` palette is used, which is quite similar to the \pkg{ColorBrewer.org}
[@color:Harrower+Brewer:2003] palette `"Purples"`. Here, only two colors are
employed, yielding a dark purple and light gray.

@@

To provide access to the HCL color palettes from within \pkg{ggplot2} graphics [@color:ggplot2; @color:ggplot2pkg]
suitable discrete and/or continuous \pkg{gglot2} color scales are provided.
The scales are named via the scheme

@@

To illustrate their usage two simple examples are shown using the qualitative `"Dark 3"`
and sequential `"Purples 3"` palettes that were also employed above. For the first example, semi-transparent
shaded densities of the sepal length from the `iris` data are shown, grouped by species
(see the left panel of Figure\ \ref{fig:iris-diamonds}).

@@

In Figure\ \ref{fig:allplots-sequential}, a heatmap (based on the well-known Maunga Whau volcano data) is used as a typical
application for a sequential palette. The elevation of the volcano is brought out clearly,
using dark colors to give emphasis to higher elevations.
The other two displays show that hue is constant in the palette while luminance and chroma vary.
Luminance increases monotonically from dark to light (as required for a proper sequential palette).
Chroma is triangular-shaped which allows to better distinguish the middle colors in the palette
when compared to a monotonic chroma trajectory.

@@

The color spaces, implemented in \pkg{colorspace}, along with their corresponding \proglang{S}4 classes
and eponymous class constructors, are:

* `RGB()` for the classic Red-Green-Blue color model that mixes three primary colors
  with different intensities to obtain a spectrum of colors. The advantage of this
  color model is (or was) that it corresponded to how computer and TV screens generated
  colors, hence it was widely adopted and still is the basis for color specifications
  in many systems. For example, the hex color codes are employed in HTML but also in \proglang{R}.
  However, the RGB model also has some important drawbacks: It does not take into
  account the output device properties, it is not perceptually uniform (a unit step
  within RGB does not produce a constant perceptual change in color), and it is
  unintuitive for humans to specify colors (say brown or pink) in this space.
  See @color:Wiki+RGB for more details.
* `sRGB()` addresses the issue of device dependency by adopting a so-called gamma
  correction. Therefore, the gamma-corrected standard RGB (sRGB), as opposed to the
  linearized RGB above, is a good model for specifying colors in software and for
  hardware. But it is still unintuitive for humans to work directly with this color space.
  Therefore, sRGB is a good place to end up in a color space manipulation but
  it is not a good place to start. See @color:Wiki+sRGB for more details.
* `HSV()` is a simple transformation of the (s)RGB space that tries to capture the
  perceptual axes: _hue_ (dominant wavelength, the type of color), _saturation_
  (colorfulness), and _value_ (brightness, i.e., light vs.\ dark). Unfortunately,
  the three axes in the HSV model are confounded so that, e.g., brightness changes
  dramaticaly with hue.
  See @color:Wiki+HSV for more details.
* `HLS()` (Hue-Lightness-Saturation) is another transformation of (s)RGB that tries
  to capture the perceptual axes. It does a somewhat better job but the dimensions
  are still strongly confounded.
  See @color:Wiki+HSV for more details.
* `XYZ()` was established by the CIE (Commission Internationale de l'Eclairage)
  based on psychophysical experiments with human subjects. It provides a unique triplet of
  XYZ values, coding the standard observer's perception of the color. It is
  device-independent but it is not perceptually uniform and the XYZ coordinates
  have no intuitive meaning.
  See @color:Wiki+CIEXYZ for more details.
* `LUV()` and `LAB()` were therefore proposed by the CIE as perceptually uniform
  color spaces where the former is typically preferred for emissive technologies
  (such as screens and monitors) whereas the latter is usually preferred when
  working with dyes and pigments. However, the three axes of these two spaces
  still do not correspond to human perceptual axes.
  See @color:Wiki+CIELUV; @color:Wiki+CIELAB for more details.
* `polarLUV()` and `polarLAB()` take polar coordinates in the UV plane
  and AB plane, respectively. Specifically, the polar coordinates of the LUV
  model are known as the HCL (Hue-Chroma-Luminance) model. These capture the
  human perceptual axes very well without confounding effects as in the HSV or
  HLS approaches. (More details follow below.)

All \proglang{S}4 classes for color spaces inherit from a virtual class `color` which is
internally always represented by matrices with three columns (corresponding to
the different three dimensions).

Note that since the inception of the color space conversion tools within \pkg{colorspace}'s
\proglang{C} code by @color:Ihaka:2003 other \proglang{R} tools for this purpose became available.
Notably, base \proglang{R} meanwhile provides `grDevices::convertColor()`
[computed in high-level \proglang{R}, @color:R] and
`farver::convert_colour()` [@color:farver] is based on a \proglang{C++} library. For many basic color
conversion purposes the \pkg{colorspace} package and these alternatives are essentially equally suitable
[see the discussion in @color:convertColor]. For more complex conversions, including
different chromatic adaptation algorithms, a more comprehensive color science approach
is implemented in the \proglang{R} package \pkg{colorscience} [@color:colorscience].
Finally, base \proglang{R} also provides
`grDevices::hcl()` for mapping HCL representations to hex codes.

@@

As motivated in the previous section,
the HCL space is particularly useful for specifying individual colors and color palettes, as its three axes match those
of the human visual system very well. Therefore, the \pkg{colorspace} package provides
three palette functions based on the HCL model: `qualitative_hcl()`, `sequential_hcl()`,
and `diverging_hcl()`. Their construction principles are exemplified in
Figure\ \ref{fig:hcl-palettes-principles} and explained in more detail below. The desaturated
palettes in the second row of Figure\ \ref{fig:hcl-palettes-principles}
bring out clearly that luminance differences (= light-dark contrasts)
are crucial for sequential and diverging palettes while qualitative palettes
are balanced at the same luminance.

@@

As suggested by @color:Ihaka:2003 `qualitative_hcl()` distinguishes the underlying categories by a sequence of
hues while keeping both chroma and luminance constant, to give each color in the
resulting palette the same perceptual weight. Thus, `h` should be a pair of
hues (or equivalently `h1` and `h2` can be used) with the starting and
ending hue of the palette. Then, an equidistant sequence between these hues is
employed, by default spanning the full color wheel (i.e., the full 360 degrees).
Chroma `c` (or equivalently `c1`) and luminance `l` (or equivalently
`l1`) are constants.

@@

When qualtitative palettes are employed for shading areas in statistical displays
(e.g., in bar plots, pie charts, or regions in maps), lighter colors (with moderate
chroma and high luminance) such as `"Pastel 1"` or `"Set 3"` are typically less
distracting. By contrast, when coloring points or lines, more flashy
colors (with high chroma) are often required: On a white background a moderate
luminance as in `"Dark 2"` or `"Dark 3"` usually works better while on a black/dark
background the luminance should be higher as in `"Set 2"`.
Some examples with demo graphics are provided in Section\ \ref{sec:palette_visualization}.

@@

As suggested by @color:Zeileis+Hornik+Murrell:2009, `sequential_hcl()` codes the underlying numeric values by a monotonic sequence
of increasing (or decreasing) luminance. Thus, the function's `l` argument should provide
a vector of length 2 with starting and ending luminance (equivalently, `l1` and
`l2` can be used). Without chroma (i.e., `c = 0`), this simply corresponds
to a grayscale palette like `gray.colors()`, see `"Grays"` and `"Light Grays"`
in Figure\ \ref{fig:hcl-palettes-sequentials}.

@@

For distinguishing colors in the center of the palette, two strategies can be employed:
(a) Hue can be varied as well by specifying an interval of hues in `h` (or beginning
hue `h1` and ending hue `h2`). More details are provided in the next section.
(b) Instead of a decreasing chroma, a triangular chroma trajectory can be employed
from `c1` over `cmax` to `c2` (equivalently specified as a vector `c` of
length 3). This yields high-chroma colors in the middle of the palette that are
more easily distinguished from the dark and light extremes. See `"Blues 3"`, `"Purples 3"`,
`"Reds 3"`, and `"Greens 3"` in Figure\ \ref{fig:hcl-palettes-sequentials}.

Instead of employing linear trajectories in the chroma or luminance coordinates,
some palettes employ a power transformation of the chroma and/or luminance
trajectory. Either a vector `power` of length 2 or separate `p1` (for chroma)
and `p2` (for luminance) can be specified. If the latter is missing, it defaults to the
former.

@@

All except the last one are inspired by the \pkg{ColorBrewer.org} palettes with
the same base name [@color:Harrower+Brewer:2003] but restricted to a single hue only.
They are intended for a white/light background. The last palette (`"Oslo"`) is taken from
the scientific color maps of @color:Crameri:2018 and is intended for a black/dark
background and hence the order is reversed starting from a light blue (not a light gray).

@@

This uses a slight power transformation with `p1 = 1.1` in the blue arm of the palette but otherwise essentially
corresponds to what `cividis_hcl()` does. For convenience `divergingx_hcl(n, palette = "Cividis")` is preregistered
using the above parameters.

@@

Next, we demonstrate a more complex example of a `swatchplot()` with three matrices
of sequential color palettes of blues, purples, reds, and greens
(see Figure\ \ref{fig:swatch-brpg}). For all palettes,
luminance increases monotonically to yield a proper sequential palette. However,
the hue and chroma handling is somewhat different to emphasize different parts of the
palette.

@@

While the `specplot()` function above works well for bringing out the HCL coordinates
associated with a given palette, it does not show how the palette fits into
the HCL space. For example, it is not so clear whether high chroma values are
close to the maximum possible for a given hue. Thus, it cannot be judged so easily
how the parameters of the hue, chroma, and luminance trajectories can be
modified to obtain another palette.

@@

```{r hcl-qualitative, fig.width = 4, fig.height = 4, dev = "png", out.width = "0.5\\textwidth", fig.cap = "Hue-chroma plane with luminance fixed at $L = 70$ along with the qualitative \\texttt{\"Dynamic\"} palette with varying hue $H$ and chroma fixed at $C = 50$."}
hclplot(qualitative_hcl(9, "Dynamic"))
```

@@

```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, dev = "png", fig.cap = "Luminance-chroma plane with variations of blue sequential single-hue palettes (similar to \\texttt{\"Blues 2\"} and \\texttt{\"Blues 3\"}). Left: Linear chroma for $H = 260$. Center: Triangular chroma for $H = 245$. Right: Power-transformed triangular chroma for $H = 245$."}
par(mfrow = c(1, 3))
hclplot(sequential_hcl(7, h = 260, c = 80, l = c(35, 95), power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95),
  power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95),
  power = c(0.8, 1.4)))
```

@@

The workhorse function to emulate color-vision deficiency is `simulate_cvd()`, which can take any vector
of valid \proglang{R} colors and transform them according to a certain CVD transformation
matrix and transformation equation. The transformation matrices have been established
by @color:Machado+Oliveira+Fernandes:2009 and are provided in objects `protanomaly_cvd`,
`deutanomaly_cvd`, and `tritanomaly_cvd`.
The convenience interfaces `deutan()`, `protan()`, and `tritan()` are the high-level functions for
simulating the corresponding kind of color blindness with a given severity (calling
`simulate_cvd()` internally).
